---
title: "AMR Gene Analysis"
author: "Shawn Higdon"
date: "2022-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Libraries

```{r, message=FALSE, include=FALSE}

packages <- c("tidyverse", "ggrepel", "ComplexHeatmap", "RColorBrewer", "viridis",
              "reshape2", "patchwork", "ggpubr", "microbiome", "microbiomeutilities",
              "phyloseq", "picante", "metacoder")

lapply(packages, library, character.only = TRUE)
```


# Read in Data

> Load AMR analytics matrix output from [STAG](https://github.com/ctmrbio/stag-mwc/blob/master/docs/source/modules.rst#amrplusplus_v2) run.

```{r}
# read in csv as df
amr.data <- read.csv("data/AMR_analytics_matrix.csv", header = T)

# extract rownames as variable
amr.data$annotation <- row.names(amr.data)

# convert to tibble
amr.df <- as_tibble(amr.data)

# split annotation by delimiter '|' with the resulting new variables 
amr.df <- amr.df |> separate("annotation",
                   into = c("lvl_5", "lvl_1", "lvl_2", "lvl_3", "lvl_4", "bonus"),
                   sep = "\\|",
                   remove = FALSE)


```


# Data Breakdown

## Annotations
```{r}
amr.df |> select('lvl_1', 'lvl_2', 'lvl_3', 'lvl_4', 'lvl_5') |> modify_if(is.character, as.factor) -> amr.anno
str(amr.anno)
summary(amr.anno)
```

# Make Phyloseq

## Counts
```{r}
# make count_matrix from amr.df
amr.otu <- amr.df |> select(1:30, 'lvl_5')
amr.otu <- as.data.frame(amr.otu)
rownames(amr.otu) <- amr.otu$lvl_5
amr.otu <- amr.otu[,-31]
amr.otu.mat <- as.matrix(amr.otu)

# phyloseq it
amr.otu.ps = otu_table(amr.otu.mat, taxa_are_rows = TRUE)
```


## Tax Table
```{r}
# make taxonomy table for amr gene hits from amr.anno
amr.tax <- as.data.frame(amr.anno)
rownames(amr.tax) <- amr.tax$lvl_5
amr.tax <- amr.tax[,-5]
amr.tax.mat <- as.matrix(amr.tax)

# phyloseq it
amr.tax.ps = tax_table(amr.tax.mat)
```


## Make Phyloseq
```{r}
amr.ps0 <- phyloseq(amr.otu.ps, amr.tax.ps)
amr.ps0
```


## Add Sample Data
```{r}
# Read in the sample map file (list has strictly static order, numeric ascending)
sample_map <- read.csv("data/sample_map_file.csv", header = T, row.names = 1, stringsAsFactors = T)

# create phyloseq sample map from dataframe
sampledata <- sample_data(sample_map, errorIfNULL = TRUE)

amr.ps <- merge_phyloseq(amr.ps0, sampledata)
amr.ps
```


# Evaluate Dataset

## Sample Sums

```{r}
sort(sample_sums(amr.ps))
```


> P1S1 was removed by STAG
> P1WF was removed by STAG
> P4B1 was removed by STAG
> P5WF has very low total counts for mapped reads >> suggest to remove from analysis since it was for the microbiome community analysis.
> P4WF has very low total counts for mapped reads >> suggest to remove from analysis since it was for the microbiome community analysis.


# Data Normalisation and Preprocessing

> The data was normalized to account for variation among samples in sequencing depth using rarefaction method implemented by STAG prior to reading results into R.
>
> Additionally, the data can be filtered to prune taxa acting as "noise" with low counts and prevalence overall
>
> These normalized counts are then transformed 2 ways:
>
> * Centered-log-ratio transformation (sparsity)
>
> * Relative Abundance
>

```{r Preprocessing bp transform}
# prune taxa (AMR genes)
amr.ps2 <- phyloseq::prune_taxa(phyloseq::taxa_sums(amr.ps) > 1000, amr.ps)
amr.ps
amr.ps2

# relative abundance
amr.ps.relab <- transform_sample_counts(amr.ps2, function(x){ x / sum(x)})

# Centered Log Ratio Transformation (CLR) transform to adjust for sparsity (zeros)
amr.ps.clr <- transform(amr.ps2, transform = 'clr')

```

> Pruning to keep only AMR genes with mapped read counts across all samples ≥ 1000 removes 986 AMR genes

# All Gene Analysis

## Beta Diversity Tests

```{r}
# Define metadata
metadata <- as(sample_data(amr.ps), "data.frame")

# calculate AMR gene composition distances
amr.clr.distmat <- phyloseq::distance(amr.ps.clr, method = "euclidean")
```


### Location

#### PERMANOVA
```{r}
#reproducible
set.seed(8888)

# calculate microbiome AMR composition distances
# ADONIS test

## location
amr.pnova_location <- adonis2(phyloseq::distance(amr.ps.clr, method="euclidean") ~ location,
                     data = metadata, method = "euclidean")
amr.pnova_location
```


#### BETA DISPERSION

```{r}

#Dispersion test and plot
amr.dispr_loc <- vegan::betadisper(amr.clr.distmat, phyloseq::sample_data(amr.ps.clr)$location)
amr.dispr_loc

#Ordination Centroids and Dispersion Labeled: Aitchison Distance
amr.ord_cen <- plot(amr.dispr_loc, main = "", sub = "")
amr.ord_cen

# extract data for ggplot
#ord_sites.long <- BiodiversityR::sites.long(ord_cen)
# boxplot
amr.ord_box_loc <- boxplot(amr.dispr_loc, main = "", xlab = "", xaxt='n')
amr.tick <- seq_along(amr.ord_box_loc$names)
axis(1, at = amr.tick, labels = FALSE)
text(amr.tick, par("usr")[2] - 0, amr.ord_box_loc$names, srt = 45, xpd = TRUE)

# permutation test
permutest(amr.dispr_loc)


```


#### ORDINATION

```{r}
# create sample_id variable
sample_data(amr.ps.clr)['sample_id'] <- row.names(sample_data(amr.ps.clr))

# set orination parameters
amr_ord_all_clr <- phyloseq::ordinate(amr.ps.clr, "RDA")

#Plot scree plot
phyloseq::plot_scree(amr_ord_all_clr) + 
  geom_bar(stat="identity", fill = "blue") +
  labs(x = "\nAxis", y = "Proportion of Variance\n")

#Scale axes and plot ordination
clr1 <- amr_ord_all_clr$CA$eig[1] / sum(amr_ord_all_clr$CA$eig)
clr2 <- amr_ord_all_clr$CA$eig[2] / sum(amr_ord_all_clr$CA$eig)

# plot ordination
amr_ord_all_clr_loc_pca <- phyloseq::plot_ordination(amr.ps.clr, amr_ord_all_clr, type = "samples", color="location")

# aesthetics of ordination plot
amr_ord_all_clr_loc_pca <- amr_ord_all_clr_loc_pca + 
  geom_point(size = 2) +
  coord_fixed(clr2 / clr1) +
  stat_ellipse(aes(group = location), linetype = 2) +
  labs(color = "Sample Location") +
  geom_text_repel(label = sample_data(amr.ps.clr)$sample_id, show.legend = FALSE) +
  theme(legend.title = element_text(size = 14),
        legend.text = element_text(face = "bold", size = 10),
        title = element_text(face = "bold", size = 14),
        axis.title.x = element_text(face = "bold", size = 13),
        axis.title.y = element_text(face = "bold", size = 13),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12))

# print ordination plot
amr_pca_allsam_loc <- print(amr_ord_all_clr_loc_pca)

amr_pca_allsam_loc

ggsave("output/plots/pca_amr_location.pdf", amr_pca_allsam_loc, dpi = 600)

```


### Location Type

##### PERMANOVA
```{r}
#reproducible
set.seed(8888)

# calculate microbiome AMR composition distances
# ADONIS test

## location
amr.pnova_ltype <- adonis2(phyloseq::distance(amr.ps.clr, method="euclidean") ~ location_type,
                     data = metadata, method = "euclidean")
amr.pnova_ltype
```

#### BETA DISPERSION

```{r}

#Dispersion test and plot
amr.dispr_lt <- vegan::betadisper(amr.clr.distmat, phyloseq::sample_data(amr.ps.clr)$location_type)
amr.dispr_lt

#Ordination Centroids and Dispersion Labeled: Aitchison Distance
amr.ord_cen <- plot(amr.dispr_lt, main = "", sub = "")
amr.ord_cen

# extract data for ggplot
#ord_sites.long <- BiodiversityR::sites.long(ord_cen)
# boxplot
amr.ord_box_lt <- boxplot(amr.dispr_lt, main = "", xlab = "", xaxt='n')
amr.tick <- seq_along(amr.ord_box_lt$names)
axis(1, at = amr.tick, labels = FALSE)
text(amr.tick, par("usr")[2] - 0, amr.ord_box_lt$names, srt = 45, xpd = TRUE)

# permutation test
permutest(amr.dispr_lt)


```


#### ORDINATION

```{r}
# create sample_id variable
sample_data(amr.ps.clr)['sample_id'] <- row.names(sample_data(amr.ps.clr))

# set orination parameters
amr_ord_all_clr <- phyloseq::ordinate(amr.ps.clr, "RDA")

#Plot scree plot
phyloseq::plot_scree(amr_ord_all_clr) + 
  geom_bar(stat="identity", fill = "blue") +
  labs(x = "\nAxis", y = "Proportion of Variance\n")

#Scale axes and plot ordination
clr1 <- amr_ord_all_clr$CA$eig[1] / sum(amr_ord_all_clr$CA$eig)
clr2 <- amr_ord_all_clr$CA$eig[2] / sum(amr_ord_all_clr$CA$eig)

# plot ordination
amr_ord_all_clr_lt_pca <- phyloseq::plot_ordination(amr.ps.clr, amr_ord_all_clr, type = "samples", color="location_type")

# aesthetics of ordination plot
amr_ord_all_clr_lt_pca <- amr_ord_all_clr_lt_pca + 
  geom_point(size = 2) +
  coord_fixed(clr2 / clr1) +
  stat_ellipse(aes(group = location_type), linetype = 2) +
  labs(color = "Sample Location Type") +
  geom_text_repel(label = sample_data(amr.ps.clr)$sample_id, show.legend = FALSE) +
  theme(legend.title = element_text(size = 14),
        legend.text = element_text(face = "bold", size = 10),
        title = element_text(face = "bold", size = 14),
        axis.title.x = element_text(face = "bold", size = 13),
        axis.title.y = element_text(face = "bold", size = 13),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12))

# print ordination plot
amr_pca_allsam_lt <- print(amr_ord_all_clr_lt_pca)

amr_pca_allsam_lt

```






# AMR Composition

```{r}

### DECIDE how many taxa you want to see
topN = 14 # all others will be grouped in "Others"

mycolors <- colorRampPalette(brewer.pal(8, "Paired"))(topN+1)
```


### By Location

#### Top L1
```{r}
amr_topL1 <- amr.ps2 %>%  #choose dataset
  psmelt %>% group_by(lvl_1) %>% # melt the table and group by tax level
  summarise(Abundance=sum(Abundance)) %>% # find most overall abundant taxa
  arrange(desc(Abundance)) %>% # order them 
  mutate(aggTaxo=as.factor(case_when( # aggTaxo will become the plot legend
    row_number()<=topN~lvl_1, row_number()>topN~'Others'))) %>% 
  dplyr::select(-Abundance) %>% #flush abundance value
  head(n=topN+1) # +1 to include the Others section! 

### CREATE an object that will be fed into ggplot
amr_comp_plot.L1 <- amr.ps2 %>% # choose dataset
  psmelt %>% # "melt" the phyloseq object into a table
  inner_join(.,amr_topL1, by="lvl_1") %>% # use topTaxa to join aggTaxo variable
  aggregate(Abundance~location+aggTaxo, # Abundance is aggregated...
            data=., FUN=sum) # ...sums "Others" taxa by specified variables

# Relative abundance counts plot
amr_relab_top_L1_location <- ggplot(amr_comp_plot.L1, aes(x=location, y=Abundance, fill=aggTaxo)) +
  geom_bar(stat="identity",position="fill") + # change to "stack" to plot total counts
  labs(title="Top L1 AMR Class",
       fill="MEGARES Level 1", x="Location", y="Relative abundance") +
  #facet_wrap(~sample_type) + # RHS of the tilde is columns; add scales="free" to hide samples without counts
  scale_fill_manual(values = mycolors,breaks=c(pull(amr_topL1, lvl_1),"Others")) +
  theme(axis.line=element_blank(),
        axis.ticks=element_blank(),
        plot.margin = unit(c(1,0.5,1,2), "cm"),
        plot.caption.position="plot",
        axis.title.y.left = element_text(vjust=3),
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1),
        axis.title.x = element_blank())

amr_relab_top_L1_location

ggsave("./output/plots/amr_relab_top_L1_location_amrplusplus.pdf", amr_relab_top_L1_location)
```

#### Top L2
```{r}
amr_topL2 <- amr.ps2 %>%  #choose dataset
  psmelt %>% group_by(lvl_2) %>% # melt the table and group by tax level
  summarise(Abundance=sum(Abundance)) %>% # find most overall abundant taxa
  arrange(desc(Abundance)) %>% # order them 
  mutate(aggTaxo=as.factor(case_when( # aggTaxo will become the plot legend
    row_number()<=topN~lvl_2, row_number()>topN~'Others'))) %>% 
  dplyr::select(-Abundance) %>% #flush abundance value
  head(n=topN+1) # +1 to include the Others section! 

### CREATE an object that will be fed into ggplot
amr_comp_plot.L2 <- amr.ps2 %>% # choose dataset
  psmelt %>% # "melt" the phyloseq object into a table
  inner_join(.,amr_topL2, by="lvl_2") %>% # use topTaxa to join aggTaxo variable
  aggregate(Abundance~location+aggTaxo, # Abundance is aggregated...
            data=., FUN=sum) # ...sums "Others" taxa by specified variables

# Relative abundance counts plot
amr_relab_top_L2_location <- ggplot(amr_comp_plot.L2, aes(x=location, y=Abundance, fill=aggTaxo)) +
  geom_bar(stat="identity",position="fill") + # change to "stack" to plot total counts
  labs(title="Top L2 AMR Class",
       fill="MEGARES Level 2", x="Location", y="Relative abundance") +
  #facet_wrap(~sample_type) + # RHS of the tilde is columns; add scales="free" to hide samples without counts
  scale_fill_manual(values = mycolors,breaks=c(pull(amr_topL2, lvl_2),"Others")) +
  theme(axis.line=element_blank(),
        axis.ticks=element_blank(),
        plot.margin = unit(c(1,0.5,1,2), "cm"),
        plot.caption.position="plot",
        axis.title.y.left = element_text(vjust=3),
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1),
        axis.title.x = element_blank())

amr_relab_top_L2_location

ggsave("./output/plots/amr_relab_top_L2_location_amrplusplus.pdf", amr_relab_top_L2_location)
```

#### Top L3
```{r}
amr_topL3 <- amr.ps2 %>%  #choose dataset
  psmelt %>% group_by(lvl_3) %>% # melt the table and group by tax level
  summarise(Abundance=sum(Abundance)) %>% # find most overall abundant taxa
  arrange(desc(Abundance)) %>% # order them 
  mutate(aggTaxo=as.factor(case_when( # aggTaxo will become the plot legend
    row_number()<=topN~lvl_3, row_number()>topN~'Others'))) %>% 
  dplyr::select(-Abundance) %>% #flush abundance value
  head(n=topN+1) # +1 to include the Others section! 

### CREATE an object that will be fed into ggplot
amr_comp_plot.L3 <- amr.ps2 %>% # choose dataset
  psmelt %>% # "melt" the phyloseq object into a table
  inner_join(.,amr_topL3, by="lvl_3") %>% # use topTaxa to join aggTaxo variable
  aggregate(Abundance~location+aggTaxo, # Abundance is aggregated...
            data=., FUN=sum) # ...sums "Others" taxa by specified variables

# Relative abundance counts plot
amr_relab_top_L3_location <- ggplot(amr_comp_plot.L3, aes(x=location, y=Abundance, fill=aggTaxo)) +
  geom_bar(stat="identity",position="fill") + # change to "stack" to plot total counts
  labs(title="Top L3 AMR Classes",
       fill="MEGARES Level 3", x="Location", y="Relative abundance") +
  #facet_wrap(~sample_type) + # RHS of the tilde is columns; add scales="free" to hide samples without counts
  scale_fill_manual(values = mycolors,breaks=c(pull(amr_topL3, lvl_3),"Others")) +
  theme(axis.line=element_blank(),
        axis.ticks=element_blank(),
        plot.margin = unit(c(1,0.5,1,2), "cm"),
        plot.caption.position="plot",
        axis.title.y.left = element_text(vjust=3),
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1),
        axis.title.x = element_blank(),
        legend.text = element_text(size = 8))

amr_relab_top_L3_location

ggsave("./output/plots/amr_relab_top_L3_location_amrplusplus.pdf", amr_relab_top_L3_location)
```


Next Steps:

[?] Correlation between ARG Subgroup Abundance and Microbial Community Abundance.
[?] Co-occurrence of MRG with ARG based on location site.

Correlate a taxonomic unit --> 

[?] MAGs

 * Highest reads and highest ARG count


Isolate all the mercury resistant?

Metal Resistance





